---
title: "STAT 184 Final Project" 
subtitle: "Preliminary EDA"
author: "Mitchell Wentzel"
output: html_notebook
---

## Front Matter

```{r}
# clean up workspace environment
rm(list = ls())

# all packages used for the assignment
library(mosaicData)
library(DataComputing)
library(tidyr)
```

# Preliminary EDA

## Research Question

How does age affect performance in major league sports careers AND do these findings vary across different major sports leagues?


## Information on Data Sources

### Primary Data:

```{r}
nhl = read.delim("NHLdata.txt", header=TRUE)
summary(nhl)
```

* Where: Data found at hockey-reference.com, derived from the National Hockey League 2019-2020 Skater Season Statistics data set.  I say derived because the original data set had almost 30 variables, so I condensed it to the 10 variables that I saw as most beneficial to my research question.

* Who: The statisticians working for the NHL and the website both collect and maintain the data from year to year, as new statistics are taken.

* Why: This data was originally collected because these same statistics are taken over the course of every NHL season and stored in this website database.

* When: My specific data were taken from the 2019-2020 NHL season.

*  What: A case in this data represents an NHL player and their attributes and statistics from the 2019-2020 NHL season.

```{r}
nhl %>%
  head(5)
```

* Some of the main variables that I plan to use in my research are AGE (Player Age), POS (Player Position), GP (Games Played), PTS (Points Scored), and TM (Team).

```{r}
names(nhl)
```

### Secondary Data:

```{r}
nfl = read.delim("NFLdata.txt", header=TRUE)
summary(nfl)
```

* Where: Data found at pro-football-reference.com, derived from the National Football League 2019 Season Scoring Leaders Statistics data set. I again say derived because the original data set had about 17 varaiables, so I condensed it to the 7 variables that I saw as most benefical to my research question and most relatable to my primary data set.

* Who: The statisticians working for the NFL and the website both collect and maintain the data from year to year, as new statistics are taken.

* Why: This data was originally collected because these same statistics are taken over the course of every NFL season and stored in this website database.

* When: My specific data were taken from the 2019 NFL season.

*  What: A case in this data represents an NFL player and their attributes and scoring statistics from the 2019 NFL season.

```{r}
nfl %>%
  head(5)
```

* Some of the main variables I plan to use in my research are Age (Player Age), Tm (Team), Pos (Player Position), G (Games Played), Pts (Points Scored) and Pts/G (Average Points per game).

```{r}
names(nfl)
```

### Explore Intuition

* Relationship between Age and Points Scored in NHL data.
```{r}
nhl %>%
  ggplot(aes(x=Age, y=PTS)) +
  geom_point()
```

* Relationship between Age and Points Scored in NFL data.
```{r}
nfl %>%
  ggplot(aes(x=Age, y=Pts)) +
  geom_point()
```

* Relationship between Age and Points Scored in NHL data, by Positions.
```{r}
nhl %>%
  ggplot(aes(x=Age, y=PTS)) +
  geom_point() +
  facet_wrap(~Pos, ncol = 3)
```

* Relationship between Age and Points Scored in NFL data, by a few positions.
```{r}
nfl %>%
  filter(Pos == c("QB", "RB", "WR")) %>%
  ggplot(aes(x=Age, y=Pts)) +
  geom_point() +
  facet_wrap(~Pos, ncol = 3)
```

* Relationship between Age and Games Played in NHL.
```{r}
nhl %>%
  ggplot(aes(x=Age, y=GP)) +
  geom_point()
```

* Relationship between Age and Games Played in NFL.
```{r}
nfl %>%
  ggplot(aes(x=Age, y=G)) +
  geom_point()
```

Above, I have created 6 basic informative plots that could help me in the start of my research. 

* In the first two plots, we can see that there is some kind of relationship between age and scoring. Even though this is good information, I will want to explore the relationships between age and other variables as well. 

* So to get a sense of some of those plots, I created the middle two plots, which show the same relationship but sorted into the facets of a couple teams and positions. 

* Finally, the last two plots show the relationship between Age and Games Played, to make sure we can see if that is a significant factor as well.

Overall, the observations from these 6 plots are very informative about some preliminary information related to my research question. This is a very good start to analyzing the data further, and diving deeper into my research question.

### Two Data Sources

* NHL 2019-2020 Data
```{r}
xfun::embed_file('NHLdata.txt')
```


* NFL 2019 Data
```{r}
xfun::embed_file('NFLdata.txt')
```

### References

* https://www.hockey-reference.com/

* https://www.pro-football-reference.com/







